---
title: "Homework 07 Multiple Linear Regression"
output:
  html_document: default
  pdf_document: default
  word_document: default
subtitle: Due by 11:59pm, Friday November 7, 2025
author: S&DS 2300/5300/ENV 757
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**1)  Model Overfitting and R-squared** 

In class, I discussed how, despite its merits, R-squared is not useful as the sole measure of predictive accuracy. The problem is that as more predictors are added (even useless ones), R-squared will always increase (assuming there is no missing data).

This problem is magnified in the case of **Model Overfitting**.   We say that a model is over-fit if the number of predictors approaches the number of observations.  When the number of predictors is one less than the number of observations, the R-squared will always be 1. 

We do a simulation to see this in action.

First, I'll simulate all of my variables (predictors and response) from a normal(0, 1) distribution. That is, the y values and ALL of the possible X predictor values are all randomly chosen from a random normal distribution with mean zero and sd = 1.  This means that the x variables are uncorrelated AND they are all NOT significant predictors of Y.


```{r}
set.seed(1)
simdata <- rnorm(10 * 15) # need 10 * 15 values simulated
simdata <- matrix(simdata, nrow = 10, ncol = 15) # now convert this vector into a matrix
colnames(simdata) <- c("y", paste0("x", 1:14)) # add column names
simdata <- as.data.frame(simdata) # convert the matrix into a data frame
```

Let's pretend that this is an actual dataset we'd like to use for linear regression, 10 observations (rows) of 14 predictors and a response variable. The columns are named as follows (so we treat the first column "y" as our response variable and the subsequent columns "x1", "x2", ... as our predictors):

```{r}
colnames(simdata)
```

**Note**: By virtue of how we simulated the data (independently and identically sampled from a standard normal distribution), "y" has no relationship with any of the 14 possible predictors.

1.1) Fit a simple linear regression model predicting `y` using `x1` and save the results to an object called `mod1`. Does it appear that `x1` is a significant predictor of `y`?  What is the value of R-squared (use `summary(mod1)$r.squared`)?  Interpret its value in the context of the model. 
 
```{r}
mod1 <- lm(y ~ x1, data = simdata)
summary(mod1)
summary(mod1)$r.squared
```


*x1 does not seem to be a significant predictor of y as the p-value is high (0.2833). Here, the R-squared is 0.1419054. In the context of the model, while x1 may explain 14.2% of the variation of y, it's most likely noise because everything was sampled iid from a normal distribution.*

1.2) Using a for-loop, now expand your linear regression model in (a) by iteratively adding in one more predictor at a time. That is to say, the first iteration of your for-loop should use `x1` as a predictor; the second iteration of your for-loop should use `x1`, and `x2`; and so on, until all 14 predictors are used in your model. Store the values of R-squared in a vector called `rsqvals` of length 14, so that the `rsqvals[i]` should contain the value of R-squared for a model using predictors `x1` through `xi`.  Finally, display the values in `rsqvals`.

*Hint:* remember the shorthand formula to include all predictors: `y ~ .`. So, for example, if I wanted to fit a model using `x1` through `x7` as predictors, I could do:

```{r}
simtemp <- simdata[,1:8]
m7 <- lm(y ~ ., data = simtemp)
```


```{r}
rsqvals <- numeric(14)
for (i in 1:14){
  simdata_subset <- simdata[, 1:(i + 1)]
  model <- lm(y ~ ., data = simdata_subset)
  rsqvals[i] <- summary(model)$r.squared
}
rsqvals
```


1.3 How many predictors did it take to reach an R-squared of 1 (write a line of code to get this value)? Display a plot that shows the increase in R-squared with an increasing number of predictors.  Be sure to label your plot.

```{r}
min(which(rsqvals == 1))

plot(1:14, rsqvals, type = "b", pch = 19, col = "violet",
     xlab = "Number of Predictors",
     ylab = "R-squared",
     main = "R-squared vs Number of Predictors")
abline(v = min(which(rsqvals == 1)), col = "blue", lty = 2)
```

*It takes 9 predictors to reach an R-squared of 1.*


**2) Ohio Crime Data** 

A survey sponsored by the US Justice Department and the University of Cincinnatti interviewed a number of Ohio residents on their attitudes toward crime, criminals, and ways of reducing crime.  In addition, various religious and demographic information was collected.

A summary of survey questions (and label information) can be found [HERE](https://github.com/jreuning/sds230_data/blob/main/OhioCrime.pdf).  You'll want to look at pages 17 - 34 at a minimum.   Each question is labeled V1, V2, etc. through V98.   We'll be looking at a subset of these questions.

The data itself is [HERE](https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/ohiocrime.csv).

2.1)  Read the data into an object called `crime`.  Get the dimension and column names of `crime`.  You won't need the `as.is = TRUE` option.

```{r}
crime <- read.csv("https://raw.githubusercontent.com/jreuning/sds230_data/refs/heads/main/ohiocrime.csv")
dim(crime)
colnames(crime)
```


2.2)  First consider the variables in columns 10 through 23; these are 14 questions having to do with attitudes toward preventing crime (see PDF file).   Each question is on a 6 point scale (see PDF file for particular levels).  Your first task is to visually examine the correlations with the `corrplot.mixed` function discussed in class 11.

I've given an outline chunk below. Your job is to fill in the details as indicated.  Be sure to remove `eval = F` before knitting.


```{r fig.height = 5, fig.width = 5}

#note the options above are to make plots work properly in the corrplot package.

#Load the corrplot package
library(corrplot)
cor1 <- cor(crime[, 10:23], use = "pairwise.complete.obs") #calculate pairwise correlations for columns 10-23 of crime.  You'll need the   use = "pairwise.complete.obs" option.

#round cor1 to 2 decimal places and display the result. 
cor1 <- round(cor1, 2)
cor1
  
#WRITE A DESCRIPTION HERE OF THE PURPOSE OF THE FOLLOWING LINE.
#The following line finds the location of the largest correlation value below 1 (therefore excluding diagonals) in cor1.
maxloc <- which(cor1 == max(cor1[cor1<1]), arr.ind = TRUE)

#WRITE A DESCRIPTION HERE OF THE PURPOSE OF THE FOLLOWING LINE.
#The following line converts the row index of the max correlation into column names of the corresponding questions, telling you which variables are involved in the max correlation
names(crime[10:23])[maxloc[1,]]

#Create an object called sigcorr that has the results of cor.mtest for columns 10-23 of the crime data.  Use 95% CI.
sigcorr <- cor.mtest(crime[, 10:23], conf.level = 0.95)

#Use corrplot.mixed to display confidence ellipses, pairwise correlation values, and put on 'X' over non-significant values.
corrplot.mixed(cor1, lower.col = "black", upper = "ellipse", tl.col = "black", number.cex = 0.7, 
               order = "hclust", tl.pos = "lt", tl.cex = 0.7, p.mat = sigcorr$p, sig.level = 0.05)
```


2.3)  Comment on the overall level of correlations among the considered questions.  Which pair of questions had the highest sample correlation?  Are you surprised? (include comments on the actual questions in your answer)

*Overall, the level of correlations is not that high among the considered questions. The pair of questions that had the highest sample correlation is V10 and V19, corresponding to the questions "During the summer, the state government should help provide jobs for inner-city youths from poor families" and "Provide job training and apprenticeship programs so that all youths will have the opportunity to get good paying jobs as adults". I am not surprised that these two would have the highest correlations among columns V10-V23, as they both are questions asking for the opinion of jobs for the youth. Additionally, both V10 and V19 use the same coding, going from 1-6 from oppose strongly to favor strongly, and having 9 = missing. Because of similarities in both question content and answer coding, it's not surprising that V10 and V19 would have the highest correlation in this subset.*

2.4) Make a scatterplot of values for the two questions that had the highest pairwise correlation.  Make sure your plot has labels for each axis (and not 'V10' - something with meaning).  Include two top titles  - one for the plot as a whole, one which reports the sample correlation to two decimal places.  How helpful is this plot?

```{r}
par(mar = c(5, 6, 4, 2))
plot(crime[, 10], crime[, 19], pch = 19, col = "red",
     xlab = "Supports state government providing jobs\n to youth from poor families",
     ylab = "Supports job training and apprenticeship\n programs for all youth",
     main = "Correlation between job related survey questions")
mtext(paste0("Correlation = ", round(cor(crime[, 10], crime[, 19], use = "pairwise.complete.obs"), 2)), line = 0)
par(mar = c(5, 4, 4, 2))
```


*This plot is not all that helpful since it is difficult to see how many responses there are for the survey questions. This makes the plot look pretty evenly spread out and a correllation = 0.58 does not make much sense visually.*

2.5)  Repeat part 2.4) but jitter results in both directions.  Write a sentence about what you observe.

```{r}
par(mar = c(5, 6, 4, 2))
plot(jitter(crime[, 10]), jitter(crime[, 19]), pch = 19, col = "red",
     xlab = "Supports state government providing jobs\n to youth from poor families",
     ylab = "Supports job training and apprenticeship\n programs for all youth",
     main = "Correlation between job related survey questions")
mtext(paste0("Correlation = ", round(cor(crime[, 10], crime[, 19], use = "pairwise.complete.obs"), 2)), line = 0)
par(mar = c(5, 4, 4, 2))
```


*This plot is a lot more helpful than the previous one, as we can not see how dense the responses are, fitting much more closely to what a correlation of 0.58 would have looked like.*

We are now going to proceed with performing stepwise regression. In particular, we're going to fit a model that looks at possible predictors of question V45 (you'll want to look up what this question is).   To do this, I'm making a new dataset called `crime2` which contains the relevant columns (notice I'm putting the response variable FIRST).  Be sure to remove the option `eval = F`.

```{r}
crime2 <- crime[, c(45, 10:23, 65, 70, 72, 87, 86)]
names(crime2)
dim(crime2)
```

2.6)  Perform best subsets regression using the `regsubsets` function in the `leaps` package.  Save the results in an object called `mod2`.  Get the summary of `mod2` and save the results in an object called `mod2sum`.  Display `mod2sum$which` to get a sense of which variables are included at each step of best subsets.

```{r}
library(leaps)
mod2 <- regsubsets(V45 ~ ., data = crime2, nvmax = ncol(crime2) - 1)
mod2sum <- summary(mod2)
mod2sum$which
```

2.7)  Following the example in classes 13-16, examine the best model according to highest r-squared.  Here are your steps:

*  Make an object called `modnum` which contains the row number in `mod2sum$which` for the model with the highest r-squared.
*  Print the variable names for predictors that ended up in this model.
*  Make a temporary dataset called `crimetemp` which has the columns of `crime2` that were included in this model.
*  Fit the model and return summary information for the model.

```{r}
modnum <- which.max(mod2sum$rsq)
names(crime2)[mod2sum$which[modnum, ]][-1]
crimetemp <- crime2[, mod2sum$which[modnum, ]]
summary(lm(V45 ~ ., data = crimetemp))
```


2.8)  Repeat 2.7) for adjusted R-squared.

```{r}
modnum <- which.max(mod2sum$adjr2)
names(crime2)[mod2sum$which[modnum, ]][-1]
crimetemp <- crime2[, mod2sum$which[modnum, ]]
summary(lm(V45 ~ ., data = crimetemp))
```


2.9)  Repeat 2.7) for BIC.

```{r}
modnum <- which.min(mod2sum$bic)
names(crime2)[mod2sum$which[modnum, ]][-1]
crimetemp <- crime2[, mod2sum$which[modnum, ]]
summary(lm(V45 ~ ., data = crimetemp))
```


2.10)  Repeat 2.7) for the Cp Statistic.

```{r}
modnum <- min(c(1:length(mod2sum$cp))[mod2sum$cp < c(1:length(mod2sum$cp))+1])
names(crime2)[mod2sum$which[modnum, ]][-1]
crimetemp <- crime2[, mod2sum$which[modnum, ]]
summary(lm(V45 ~ ., data = crimetemp))
```


2.11)  We choose as our final model the model indicated by BIC.  Go refit this model and save the results in an object called `modfin`.  How many observations had missing values in this model vs. the number of observations with missing values in the model with all predictors?

```{r}
modnum_fin <- which.min(mod2sum$bic)
names(crime2)[mod2sum$which[modnum_fin, ]][-1]
crimetemp_fin <- crime2[, mod2sum$which[modnum_fin, ]]
modfin <- lm(V45 ~ ., data = crimetemp_fin)
summary(modfin)

mod_full <- lm(V45 ~ ., data = crime2)
summary(mod_full)
```


*The model indicated by BIC had 24 missing values while the model with all predictors had 74 missing values.*

2.12)  Make two residual plots - studentized residuals vs fitted values (with boundaries at +/-2 and +/-3) as well a normal quantile plot of the residuals. Write a few sentences about how the plots do/do not indicate that we've met the assumptions of our regression model.  Note : the six-toed beast that seems to have slashed the plot of fits vs. residuals is exactly what we would expect.   Why is this?

```{r}
# source("https://raw.githubusercontent.com/jreuning/sds230_files/refs/heads/main/regJDRS.txt")
# myResPlots
library(car)
myResPlots <- function(model, label){
  #plot of fitted vs. studentized residuals
  plot(rstudent(model) ~ model$fitted.values, pch = 19, col = 'red', 
       xlab = "Fitted Values", ylab = "Studentized Residuals", 
       main = paste("Fits vs. Studentized Residuals,", label))
  abline(h = 0, lwd = 3)
  abline(h = c(2,-2), lty = 2, lwd = 2, col="blue")
  abline(h = c(3,-3), lty = 2, lwd = 2, col="green")
  
  #Normal quantile plot of studentized residuals
  qqp(rstudent(model), pch = 19, main = paste("NQ Plot of Studentized Residuals,", label))
  
  #Plot of Cooks Distances
  # plot(model, which = 4, lwd = 3, col = 'red')
  # abline(h = 6/length(model$residuals), col = 'black', lwd = 3, lty = 2)
}

myResPlots(modfin, "Model for question V45")
```


*I would say that the normal quantile plot shows that the data seems to be normal. The studentized residuals vs fitted values plot show diagonal lines going to the bottom right with the residuals generally staying with +/-2. Because of this, I would think we are not violating linearity or homoscedasticity. As to why there seem to be 6 distinct lines in the studentized residuals vs fitted values plot, this is because we are using an ordinal scale.*

2.13)  Run the code below.  Describe what this does in not more than two sentences.

```{r}

CIs <- confint(modfin)
coefs <- coef(modfin)
library(plotrix)
plotCI(1:(length(coefs)-1), coefs[-1], 
       ui = CIs[-1,2], li = CIs[-1,1], 
       xlab = "", 
       ylab = "Coefficient (and 95% CI)", 
       main = "Final Crime Model Coefficients and CI's", 
       axes = FALSE, lwd = 2, col = "blue")
abline(h = 0, lty = 2, lwd = 3, col = "red")
axis(side = 2)
axis(side = 1, at = 1:(length(coefs)-1), label = names(coefs)[-1], las = 2)

```

*The code plots the coefficients of modfin along with their 95% confidence intervals. The red line is to help illustrate how far away the coefficients are away from zero.*

2.14)  FINALLY - write a short paragraph discussing your model results.

*  Comment on R-squared
*  Comment on direction and interpretation of each of the predictors included in the final model.   Are you surprised in any instance?

*The R-squared of the final model is 0.2264, which to me does not seem to be a lot. As for direction and interpretation of each of the predictors included in the final model, I'd say all of them make sense to me. A higher score in V45 means that the individual believes more strongly that punishment is needed to prevent future crime. Higher scores in V17, V20, and V22 seem to support harsh punishments for criminals, explaining the positive coefficients. For V72, the lower the score the lower level of education, and I personally believe that being more educated would allow people to see that punishment is not the only way for people to turn away from crime, so a negative coefficient is not surprising for me. Lastly, lower scores in V10 show lower support for jobs for inner-city youths from poor families, so the negative coefficient makes sense to me since people who responded highly to V10 probably believe that early financial opportunity can curb crime rate.*

3)  Pseudocode

Prompt:
You’ve been hired by the Connecticut Department of Environmental Protection to help understand what drives nutrient pollution in lakes across the state. The dataset you’re given includes measurements for total phosphorus concentration (your response variable) and nineteen potential predictors, including lake area, depth, surrounding land use, population density, road density, and several management practices. Your boss wants you to determine which combination of predictors best explains variation in phosphorus levels, while also checking that your final model meets the assumptions of linear regression. Write pseudocode describing the steps you would take to provide meaningful insights into how different variables contribute to better or worse water quality (based on total phosphorus concentration) and what these variables may suggest about future protection, clean up, and management decisions (i.e. outline the steps you'd use to fit a multiple linear regression model).

- *Load in data*
- *Do multiple pairwise correlations for the potential predictors*
- *Identify which predictors are correlated to inform future policy making*
- *Perform best subsets regression according to highest r-squared using the regsubsets function. Repeat this for adjusted R-squared, BIC, and Cp Statistic.*
- *Choose a final model*
- *Make a studentized residuals vs fitted values and a normal quantile plot of the residuals. This is to ensure the assumptions of normality, linearity, and homoscedasticity. If these assumptions are broken, choose a different model that was generated before.*
- *Make a plot showing the model's coefficients and CIs to inform policy decisions (i.e., which coefficients are far away from 0 and what direction are they in.*

