---
title: "expanded_query_results"
output: html_document
---

```{r}
results <- read.csv("/Users/andrewyu/Documents/Yale/Fall 2025/Clinical Database Management Systems and Ontologies/Final Project/expanded_query_results_v3.csv")
colnames(results)
dim(results)
```

```{r}
length(unique(results$subject_id))
length(unique(results$hadm_id))
length(unique(results$stay_id))
```

```{r}
boxplot(min_pf_ratio ~ inhospital_death,
        data = results,
        xlab = "In-Hospital Death",
        ylab = "Minimum PaO2/FiO2 Ratio",
        names = c("No", "Yes"),
        col = c("skyblue", "pink"),
        main = "PF Ratio by In-Hospital Death")
```
According to (https://litfl.com/pao2-fio2-ratio/), at sea level, the normal PaO2/FiO2 ratio is ~ 400-500 mmHg (~55-65 kPa).
```{r}
# Group by inhospital_death
bp <- boxplot(min_pf_ratio ~ inhospital_death, data = results, plot = FALSE)

# bp$out contains all outlier values
length(bp$out)  # total number of outlier values
```
Here, we see that only 487 out of our 7618 min_pf_ratio readings (per each stay_id) are considered outliers, but all these patients here have sepsis. Below, we graph a more close up view just to take a look at things.

```{r}
boxplot(min_pf_ratio ~ inhospital_death,
        data = results,
        xlab = "In-Hospital Death",
        ylab = "Minimum PaO2/FiO2 Ratio",
        names = c("No", "Yes"),
        col = c("skyblue", "pink"),
        main = "PF Ratio by In-Hospital Death",
        ylim = c(0, 600))
```
In the code chunk below, we're making a table to visualize the number/percent of each people in each ards category, as well as seeing the in hospital death rate
```{r}
# Compute min PF ratio per hadm_id
min_pf_per_hadm <- aggregate(min_pf_ratio ~ hadm_id, data = results, FUN = min)

# Assign ARDS severity per admission based on min PF ratio
min_pf_per_hadm$ards_severity <- cut(
  min_pf_per_hadm$min_pf_ratio,
  breaks = c(-Inf, 100, 200, 300, Inf),
  labels = c("Severe", "Moderate", "Mild", "No ARDS")
)

# Add in-hospital death (one row per hadm_id)
hadm_death <- unique(results[, c("hadm_id", "inhospital_death")])
results_adm <- merge(min_pf_per_hadm, hadm_death, by = "hadm_id")

# Total number of admissions
N <- nrow(results_adm)

# Counts per severity
counts <- table(results_adm$ards_severity)

# Percent of total
pct <- prop.table(counts) * 100

# In-hospital death percent per severity
death_pct <- tapply(results_adm$inhospital_death, results_adm$ards_severity,
                    function(x) mean(x == 1, na.rm = TRUE) * 100)

# Assemble final summary table
summary_table <- data.frame(
  ards_severity = names(counts),
  total_admissions = paste0(counts, " (", sprintf("%.1f", pct), "%)"),
  inhospital_death_pct = sprintf("%.1f%%", death_pct),
  row.names = NULL
)

summary_table
```
Note to self: In the above table, the ards_severity is much worse than my previous query. The reason for this is becuase the hadm_id before only took one pf_ratio, but here we are getting the worst value recorded.

Here, I want to see the survival rate of each type of ICU care unit
```{r}
# 1. Aggregate stays per hadm_id
hadm_summary <- aggregate(
  first_careunit ~ hadm_id + inhospital_death,
  data = results,
  FUN = function(x) {
    u <- unique(x)
    if(length(u) > 1) "Multiple/Mixed ICU Stays" else u
  }
)

# 2. Create table for barplot
plot_table <- table(hadm_summary$first_careunit, hadm_summary$inhospital_death)
colnames(plot_table) <- c("Survived", "Died")  # 0 = Survived, 1 = Died

# 3. Order ICU types by total admissions (most to least)
total_per_icu <- rowSums(plot_table)
plot_table <- plot_table[order(total_per_icu, decreasing = FALSE), ]

# 4. Only annotate counts for top 5 bars
top_n <- 12

# 5. Expand left margin and shrink y-axis labels
par(mar = c(5, 12, 4, 5), cex.axis = 0.5)  # cex.axis < 1 shrinks labels

# 6. Horizontal stacked barplot
bp <- barplot(
  t(plot_table),      # transpose for stacking by death status
  beside = FALSE,     
  col = c("blue", "red"),
  horiz = TRUE,
  las = 1,            
  xlab = "Number of Admissions",
  main = "Admissions by First Care Unit and In-Hospital Death",
  xlim = c(0, max(colSums(plot_table)) * 1.25) # extend for survival %
)

# 7. Add counts inside bars for top 5 ICU types
for(i in seq_len(nrow(plot_table))) {
  survived <- plot_table[i, "Survived"]
  died <- plot_table[i, "Died"]
  
  y_pos <- bp[i]
  
  if(i >= top_n) {  # only for top 5
    if(survived > 0) text(survived / 2, y_pos, survived, col = "white")
    if(died > 0) text(survived + died / 2, y_pos, died, col = "white")
  }
  
  # Add survival percentage on the right for all bars
  total <- survived + died
  surv_pct <- round(survived / total * 100, 1)
  text(total + max(colSums(plot_table)) * 0.02, y_pos, paste0(surv_pct, "%"), adj = 0)
}

# 8. Legend
legend("bottomright", legend = c("Survived", "Died"), fill = c("blue", "red"))

```

Making a table of ventilation survival rate
```{r}
# # 1. Collapse ventilation per hadm_id
# vent_status <- aggregate(ventilation_status ~ hadm_id, data = results, FUN = function(x) {
#   x <- unique(x)
#   
#   if (all(x == "None")) {
#     return("No")
#   } else if (all(x != "None" & x != "")) {
#     return("Yes")
#   } else if (all(x == "")) {
#     return("Null")
#   } else {
#     return("Mixed")
#   }
# })
# 
# # 2. Merge with results_adm (one row per hadm_id)
# results_adm <- merge(results_adm, vent_status, by = "hadm_id", all.x = TRUE)
# 
# # Total number of admissions
# N <- nrow(results_adm)
# 
# # Counts per severity
# counts <- table(results_adm$ards_severity, results_adm$ventilation_status)
# 
# # Percent of total
# pct <- prop.table(counts) * 100
# 
# # In-hospital death percent per severity
# death_pct <- tapply(results_adm$inhospital_death, list(results_adm$ards_severity, results_adm),
#                     function(x) mean(x == 1, na.rm = TRUE) * 100)

# 1. Collapse ventilation per hadm_id
vent_status <- aggregate(ventilation_status ~ hadm_id, data = results, FUN = function(x) {
  x <- unique(x)
  
  if (all(x == "None")) {
    return("No")
  } else if (all(x != "None" & x != "")) {
    return("Yes")
  } else if (all(x == "")) {
    return("Null")
  } else {
    return("Mixed")
  }
})

# 2. Merge with results_adm (one row per hadm_id)
results_adm <- merge(results_adm, vent_status, by = "hadm_id", all.x = TRUE)

# 3. Total number of admissions
N <- nrow(results_adm)

# 4. Counts per severity and ventilation status
counts <- table(results_adm$ards_severity, results_adm$ventilation_status)

# 5. Percent of total
pct <- prop.table(counts) * 100

# 6. In-hospital death percent per severity and ventilation
death_pct <- tapply(
  results_adm$inhospital_death, 
  list(results_adm$ards_severity, results_adm$ventilation_status),
  function(x) mean(x == 1, na.rm = TRUE) * 100
)

# 7. Convert to data frame
summary_table <- as.data.frame(as.table(death_pct))
colnames(summary_table) <- c("ards_severity", "ventilation_status", "inhospital_death_pct")

# 8. Add counts and percent of total
summary_table$total_admissions <- paste0(
  as.vector(counts[cbind(as.character(summary_table$ards_severity), 
                         as.character(summary_table$ventilation_status))]),
  " (",
  sprintf("%.1f", as.vector(pct[cbind(as.character(summary_table$ards_severity), 
                                      as.character(summary_table$ventilation_status))])),
  "%)"
)

# 9. Format death percent
summary_table$inhospital_death_pct <- sprintf("%.1f%%", summary_table$inhospital_death_pct)

# 10. Order factors
severity_order <- c("Severe", "Moderate", "Mild", "No ARDS")
vent_order <- c("Yes", "No", "Mixed")

summary_table$ards_severity <- factor(summary_table$ards_severity, levels = severity_order)
summary_table$ventilation_status <- factor(summary_table$ventilation_status, levels = vent_order)

# 11. Order the table
summary_table <- summary_table[order(summary_table$ards_severity, summary_table$ventilation_status), ]

summary_table

```


